Cartograms with Dynamic Features

### 2022-03-04 paper/section/data.tex
draft: introduction
draft: update and review

Obtaining the heterogenous data can be challenging, especially when an Electronic Health Records (EHR) dataset is involved \cite{wang2021EHRa}. The data acquisition involves multiple steps. The first step is to obtain both geospatial boundaries and EHR data that match one another. The second step is to pre-process the EHR data to remove empty and erroneous values. The final step is to transform the data into a format that is suitable for visualization.

@@

Geospatial boundaries, or shapefiles, were obtained from the following sources. After acquiring the shapefiles, we used QGIS \cite{qgisWelcome} to manually adjust projections and merge them into one shapefile. Finally, mapshaper \cite{blochMapshaper} is used to convert the merged shapefile into a TopoJSON \cite{TopoJSON} file to reduce the file size and improve the performance of our implementation.

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=0.6\columnwidth]{figure/ccg.png}
    \caption{A map of 135 CCGs in England as of 2020, obtained from the Open Geography portalx \cite{opengeographyportalxOpen}.}
    \label{fig:ccg}
\end{figure}
}

\subsubsection{Clinical Commissioning Groups Shapefile}

Clinical Commissioning Groups (CCGs) are the main geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}. The number of CCGs changes over time due to the NHS re-organization. The most up-to-date shapefile is available from the Open Geography portalx \cite{opengeographyportalxOpen}. We decided to use the CCG shapefile from 2020 at the time of writing, because there is no public EHR data published based on the latest CCG re-organization that took place in 2021.

\subsubsection{River Features}

@@

>>>>

Obtaining the heterogenous data can be challenging, especially when an EHR dataset is involved \cite{wang2021EHRa}. The data acquisition involves multiple steps. The first step is to obtain both geospatial boundaries and EHR data that match one another. The second step is to pre-process the EHR data to remove empty and erroneous values. The final step is to transform the data into a format that is suitable for visualization.

@@

Geospatial boundaries, or shapefiles, were obtained from the sources described below.
{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=0.6\columnwidth]{figure/ccg.png}
    \caption{A map of 135 CCGs in England as of 2020, obtained from the Open Geography portalx \cite{opengeographyportalxOpen}.}
    \label{fig:ccg}
\end{figure}
}

\bobgraph{Clinical Commissioning Groups Shapefile}

Clinical Commissioning Groups (CCGs) are the main geographic unit of the National Health Service (NHS) in the UK \cite{nhsNHS}. The number of CCGs changes over time due to the NHS re-organization. The most up-to-date shapefile is available from the Open Geography portalx \cite{opengeographyportalxOpen}. We decided to use the CCG shapefile from 2020 at the time of writing, because there is still no public EHR data published based on the latest CCG re-organization that took place in 2021.

\bobgraph{River Features}

@@

After acquiring the shapefiles, we used QGIS \cite{qgisWelcome} to manually adjust projections and scales, and merge them into one shapefile. Finally, mapshaper \cite{blochMapshaper} is used to convert the merged shapefile into a TopoJSON \cite{TopoJSON} file to reduce the file size and improve the performance of our implementation.

### 2022-03-04 paper/section/introduction.tex
draft: update and review

\new{Cartograms are representations of geographical space generated by the value-by-area mapping technique that combines statistical and geographical information. Various styles of cartograms have been proposed and implemented in fields such as geography, social science, and visualization \cite{dent2009Cartography}.}

\new{In this paper, we follow the categorization of cartograms by \citea{nusrat2016State}: contiguous, non-contiguous, rectangular, and Dorling. Among these four types, trade-offs between accuracies are made (See \tableref{table:accuracy}).}

\new{
    Based on Demers cartograms, we include dynamic topological features, such as rivers and mountains, aiming to improve the topological accuracy without sacrificing statistical accuracy. We then introduce a cartographic layout algorithm that uses these topological features to update the layout of the nodes representing geographical regions. To minimize cartographic errors, the algorithm also updates the position of rivers to accommodate the node layout. We then visualize a real world Electronic Health Records (EHR) dataset to assess the accuracy of our algorithm.
}

>>>>

Cartograms are representations of geographical space generated by a value-by-area mapping technique that combines statistical and geographical information \cite{dent2009Cartography}. \new{Various styles of cartograms have been proposed and implemented, covering topics on urban planning \cite{harris2018Mapping, arranz-lopez2021Enduser}, natural hazard forecasting \cite{pappenberger2019Cartograms, park2020Flood}, conservation and environmental planning \cite{galluzzi2018Mapping, rocchini2019Cartogramming}, political and social demographics \cite{breitzman2018Using, alieva2021How}, and public health decision making \cite{gao2020Visualising, sack2021Visualizing}.}

In this paper, we follow the categorization of cartograms by \citea{nusrat2016State}: contiguous, non-contiguous, rectangular, and Dorling. Among these four types, trade-offs between accuracies are made (See \tableref{table:accuracy}). Building on Demers cartograms \cite{ian2002Cartogram}, we develop dynamic topological features, such as rivers, aiming to improve the readability and geographical accuracy without sacrificing statistical accuracy. We incorporate a cartographic layout algorithm that includes these topological features to update the layout of the nodes representing geographical regions. To minimize geographical errors, the algorithm also updates the position of rivers to accommodate the node layout. We then apply the algorithm a real world Electronic Health Records (EHR) dataset to assess the accuracy of our algorithm.

\new{Our contributions include:}

\begin{itemize}
    \item A new variant of Demers cartograms that incorporates dynamic topological features to improve readability and recognizability.
    \item A novel layout algorithm that preserves node positions relative to dynamic topological features such as rivers.
\end{itemize}

### 2022-03-04 paper/section/main.tex
draft: introduction
draft: update and review

\algoref{alg:UpdateNodePosition} provides an overview of our algorithm. We first load and render the CCG geospatial boundaries. For each CCG we compute the centroid and represent it using a square node with $ s $ as its size. We then apply the Fast Node Overlap Removal (VPSC) algorithm \cite{dwyer2006fast} repeatedly to remove overlaps. We chose VPSC over other node overlap removal algorithms since VPSC is able to provide spread minimization and node movement minimization while maintaining a good level of global shape preservation \cite{chen2020Node}. During each iteration, we compute node trajectories (See \algoref{alg:check river crossing}) and translate nodes to their new position. Nodes that cross a river are translated back to their previous position. This procedure ends when 1) no overlaps are present; 2) no nodes cross a river. We then increase $ \nodeSize $ by one pixel and repeat the algorithm until the max node size $ \nodeSizeMax $ is reached.

@@

\begin{algorithm}[tb!]
    \caption{Procedure to update node positions by removing overlap and prevent nodes from crossing rivers.}\label{alg:UpdateNodePosition}
    \textbf{Global variables:} \\
    $ \nodeList \gets $ a list of nodes representing CCGs \\
    $ \nodeSize \gets $ the current size of all nodes \\
    $ \nodeSizeMax \gets $ the maximum size of a node \\
    $ \stalemate \gets $ the maximum number of iterations indicating a stalemate \\

    \textbf{Local variables:} \\
    $ \nodeListVPSC \gets $ a list of nodes representing CCGs after running VPSC \\
    $ \crossRiver \gets $ boolean indicating a node crossed a river \\

@@

                \If {TranslateNode ($ \nodeListEach, \nodeListVPSCEach $)}
                    \State $ \crossRiver \gets $ \Call{TestRiverCross}{$ \nodeListEach, \nodeListVPSCEach $}

                    \If{$ \crossRiver = True $}
                        \State $ \stalemate \gets \stalemate + 1 $

                        \If{$ \stalemate < \stalemate $}
                            \State \Call{TranslateNode}{$ \nodeListEach $}
                        \Else
                            \State \Call{DeriveCorridor}{$ \nodeListEach, \nodeListVPSCEach $}
                        \EndIf

@@

    \textbf{Input:} \\
    $ \node \gets $ required, the node \\
    $ \nodeVPSC \gets $ optional, the node after running VPSC \\

    \textbf{Output:} \\
    A boolean for whether the node's position is updated. \\

    \begin{algorithmic}[1]
        \Procedure{TranslateNode}{$ \node, \nodeVPSC $}
        \If{$ \node.x = \nodeVPSC.x $ and $ \node.y = \nodeVPSC.y $}
            \item[] \Comment{Node position remains unchanged}
            \State \Return{$ False $}
        \EndIf

        \If{$ \nodeVPSC$ is undefined } 
        \item[] \Comment{When optional parameter $ \nodeVPSC $ is undefined}
        \item[] \Comment{Move the node back to its previous position}
            \State $ \node.x \gets \node.previous.x,~\node.y \gets \node.previous.y $
        \Else
            \State $ \node.x \gets \nodeVPSC.x,~\node.y \gets \nodeVPSC.y $
            \State $ \node.history.push(\node.x, \node.y) $
        \EndIf
        \State \Return{$ True $}
        \EndProcedure
    \end{algorithmic}
\end{algorithm}

@@

\subsection{River Cross Testing}

We use rivers as topological boundaries and prevent nodes from crossing them. When a node's position changes, we test if the node's trajectory intersects any segment of a river. See \algoref{alg:check river crossing}. A bounding box intersection test can be performed to reduce the number of edge intersection detections required.

% \begin{noindent}
\begin{algorithm}[tb!]
    \caption{Procedure to test if a node crosses a river.}\label{alg:check river crossing}
    \textbf{Input:} \\
    $ \node \gets $ required, the node \\
    $ \nodeVPSC \gets $ required, the node after running VPSC \\

    \textbf{Output:} \\
    A boolean indicating the node crosses a river. \\

@@

    \textbf{Local variables:} \\
    $ \riverEdgeListEach \gets $ a river edge in $ \riverEdgeList $ \\ 
    $ \nodeBoundingBox, \riverEdgeBoundingBoxEach \gets $ the bounding boxes for $ \node $ and $ \riverEdgeListEach $ \\
    $ \boundingBoxInt \gets $ boolean indicating $ \nodeBoundingBox $ intersects $ \riverEdgeBoundingBoxEach $ \\
    $ \nodeEdge \gets $ the edge for $ \node $\\
    $ \edgeBoxInt \gets $ boolean indicating $ \nodeEdge $ intersects $ \riverEdgeListEach $ \\

    \begin{algorithmic}[1]
        \Procedure{TestRiverCross}{$ \node, \nodeVPSC $}

        \ForEach{$ \riverEdgeListEach \in \riverEdgeList $}
            \State $ \nodeBoundingBox \gets $ GenerateBoundingBox ($ \node, \nodeVPSC $)
            \State $ \riverEdgeBoundingBoxEach \gets $ GenerateBoundingBox ($ \riverEdgeListEach $)
            \State $ \boundingBoxInt \gets $ TestBoundingIntersection ($ \nodeBoundingBox, \riverEdgeBoundingBoxEach $)

            \If{$ \boundingBoxInt = True $}
                \State $ e_{n} \gets $ GenerateEdge ($ \node, \nodeVPSC $)
                \State $ \edgeBoxInt \gets $ TestEdgeIntersection ($ \nodeEdge, \riverEdgeListEach $)

                \If{$ \edgeBoxInt = True $}
                    \State \Return{$ True $}
                \EndIf

@@

\subsection{Corridor Derivation}

As the VPSC always tries to produce an optimal node layout where node distribution and translation are minimized, a node might be repeatedly translated between two positions due to river crossing and congestion, creating a stalemate situation, as shown in \figref{fig:stalemate}. In order to create space for VPSC to generate a new layout, we introduce a parameter, $ \stalemate $, to test for stalemate. If a node is translated between two positions for more than $ \stalemate $ iterations, it is considered to be in a stalemate. A stalemate triggers a corridor derivation which moves all nodes within the corridor by $ d $ set by the user (See \figref{fig:corridor} and \algoref{alg:derive corridor}).

@@

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/corridor.png}
    \caption{When a stalemate occurs, a corridor (black rectangular) is derived based on the current (A) and previous (B) positions of the node that crossed a river, as described in \algoref{alg:derive corridor}. All nodes within the corridor are moved by $ d $ in the direction of $ \Vector{B}{A} $. }
    \label{fig:corridor}
\end{figure}
}

@@

    \textbf{Local variables:} \\
    $ \PointP \gets $ the point extending $ \Vector{\node.previous}{~\node} $ such that the length of $ \Vector{\node.previous}{~\PointP} $ is $ \CorridorLength $\\
    $ \EdgeParallelA, \EdgeParallelB \gets $ the edges parallel to $ \Vector{\node}{\PointP} $ \\
    $ corridor \gets $ a rectangular formed by $ \EdgeParallelA $ and $ \EdgeParallelB $ \\

    \begin{algorithmic}[1]
        \Procedure{DeriveCorridor}{$ \node $}
            \State TranslateNode ($ \node $)

            \State $ \PointP \gets $ \Call{DerivePoint}{$ \Vector{\node.previous}{~\node} $, $ \CorridorLength $}

            \State $ \EdgeParallelA \gets $ \Call{DeriveParallelEdge}{$ \Vector{\node.previous}{~\PointP} $, $ \frac{\CorridorWidth}{2} $}

            \State $ \EdgeParallelB \gets $ \Call{DeriveParallelEdge}{$ \Vector{\node.previous}{~\PointP} $, $ -\frac{\CorridorWidth}{2} $}

@@

            \ForEach{$ \nodeInCorridorEach $ inside $ corridor $}
                \State $ \nodeInCorridorEachDx \gets \nodeInCorridorEach.x - \nodeInCorridorEach.previous.x $

@@

    \begin{algorithm}[tb!]
        \caption{Procedure to derive a point based an edge and a distance.}\label{alg:derive corridor point}
        \textbf{Input:} \\
        $ \Edge \gets $ the edge used to derive the new point \\
        $ \Distance \gets $ the distance between $ \PointP $ and $ \Edge.start $ \\

        \textbf{Output:} \\
        A point, $ \PointP $, that is distance $ \Distance $ away from $ \Edge.start $. \\

        \textbf{Local variables:} \\
        $ \dx, \dy \gets $ the differences in $ x, y $ for $ \Edge.start $ and $ \Edge.end $ \\

        \begin{algorithmic}[1]
            \Procedure{DerivePoint}{$ \Edge $, $ \Distance $}
                \State $ \dx \gets \Edge.start.x - \Edge.end.x $

                \State $ \dy \gets \Edge.start.y - \Edge.end.y $

@@

    \textbf{Local variables:} \\
    $ \dx, \dy \gets $ the differences in $ x, y $ for $ \Edge.start $ and $ \Edge.end $ \\
    $ \Scale \gets $ the scale of $ \frac{\Distance}{\sqrt{\dx^2 + \dy^2}} $ \\

    \begin{algorithmic}[1]
        \Procedure{DeriveParallelEdge}{$ \Edge $, $ \Distance $}
            \State $ \dx \gets \Edge.start.x - \Edge.end.x $

            \State $ \dy \gets \Edge.start.y - \Edge.end.y $

@@

            \State $ \EdgeParallel.start.x \gets \Scale \cdot -\dy + \Edge.start.x $

            \State $ \EdgeParallel.start.y \gets \Scale \cdot \dx + \Edge.start.y $

            \State $ \EdgeParallel.end.x \gets \Scale \cdot -\dy + \Edge.end.x $

            \State $ \EdgeParallel.end.y \gets \Scale \cdot \dx + \Edge.end.y $

>>>>

{
\begin{figure}[b!]
    \centering
    \includegraphics[width=\columnwidth]{figure/flowchart.png}
    \caption{A flowchart abstraction of our layout algorithm.}
    \label{fig:flowchart}
\end{figure}
}

\algoref{alg:UpdateNodePosition} and \figref{fig:flowchart} provide an overview of our algorithm. \textbf{Initialization:} We first load and render the CCG geospatial boundaries. For each CCG we compute the centroid and represent it using a small square node with $ \nodeSize $ as the initial size. \textbf{Node layout:} The basic algorithm is as follows: we first apply the Fast Node Overlap Removal (VPSC) algorithm \cite{dwyer2006fast} repeatedly to remove overlaps. We chose VPSC over other node overlap removal algorithms since VPSC is able to provide spread minimization and node movement minimization while maintaining a good level of global shape preservation \cite{chen2020Node}. During each iteration, we compute node trajectories (See \algoref{alg:check river intersection}) and translate nodes to their new position. Nodes that cross a river are translated back to their previous position. This procedure ends when 1) no node overlap is present; 2) no nodes cross a river. We then increase $ \nodeSize $ by one pixel and repeat the algorithm until the max node size $ \nodeSizeMax $ is reached. \new{The size growing process provides a stability in the layout and minimizes geographical errors.}

@@

\begin{algorithm}[tb!]
    \caption{Procedure to update node positions by removing overlap and prevent nodes from crossing rivers.}\label{alg:UpdateNodePosition}
    \textbf{Global variables:} \\
    $ \nodeList \gets $ a list of nodes representing CCGs \\
    $ \nodeSize \gets $ the current size of all nodes \\
    $ \nodeSizeMax \gets $ the maximum size of a node \\
    $ \stalemateMax \gets $ the maximum number of iterations indicating a stalemate \\

    \textbf{Local variables:} \\
    $ \nodeListVPSC \gets $ a list of nodes representing CCGs after VPSC \\
    $ \crossRiver \gets $ boolean indicating a node crossed a river \\
    $ \nodeListEach \gets $ a node in $ \nodeList $ \\
    $ \nodeListEachPrevious \gets $ the previous position of $ \nodeListEach $\\
    $ \nodeListVPSCEach \gets $ a node in $ \nodeListVPSC $, which is the next position of $ \nodeListEach $ after VPSC \\

@@

                \If {\Call{TranslateNode}{$ \nodeListEach, \nodeListVPSCEach $}}
                    \State $ \crossRiver \gets $ \Call{TestRiverIntersection}{$ \nodeListEach, \nodeListVPSCEach $}

                    \If{$ \crossRiver = True $}
                        \State $ \nodeListEachStalemate \gets \nodeListEachStalemate + 1 $

                        \If{$ \nodeListEachStalemate < \stalemateMax $}
                            \State $ \nodeListEach(x,y) \gets \nodeListEachPrevious(x,y) $
                        \Else
                            \State \Call{ProcessStalemate}{$ \nodeListEach, \nodeListVPSCEach $}
                            \State $ \nodeListEachStalemate \gets 0 $
                        \EndIf

@@

    \textbf{Input:} \\
    $ \node \gets $ the node \\
    $ \nodeVPSC \gets $ the node after running VPSC \\

    \textbf{Output:} \\
    Returns $ True $ if the node's position is updated. \\

    \begin{algorithmic}[1]
        \Procedure{TranslateNode}{$ \node, \nodeVPSC $}
        \If{$ \node(x,y) = \nodeVPSC(x,y) $}
            \item[] \Comment{Node position remains unchanged}
            \State \Return{$ False $}
        \EndIf

        \State $ \node(x,y) \gets \nodeVPSC(x,y) $

        \State \Return{$ True $}
        \EndProcedure
    \end{algorithmic}
\end{algorithm}

@@

\subsection{River Intersection Testing}

We use rivers as topological boundaries and prevent nodes from crossing them. \new{The resolution of rivers can be dynamically adjusted, as shown in \figref{fig:river resolution}}. When a node's position changes, we test if the node's trajectory intersects any segment of a river. See \algoref{alg:check river intersection}. A bounding box intersection test can be performed to reduce the number of edge intersection detections required.

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/river_resolution.png}
    \caption{The resolution of rivers can be dynamically adjusted. The top figure shows River Thames at its original resolution. The bottom figure shows the river with a 60\% reduction in resolution. The reduced resolution preserves the majority of River Thames' original shape and improves the performance of our river intersection tests.}
    \label{fig:river resolution}
\end{figure}
}

% \begin{noindent}
\begin{algorithm}[tb!]
    \caption{Procedure to test if a node, $ \node $ crosses a river.}\label{alg:check river intersection}
    \textbf{Input:} \\
    $ \node \gets $ the node \\
    $ \nodeVPSC \gets $ the node after running VPSC \\

    \textbf{Output:} \\
    Returns $ True $ if the node crosses a river. \\

@@

    \textbf{Local variables:} \\
    $ \nodeBoundingBox, \riverEdgeBoundingBoxEach \gets $ the bounding boxes for $ \nodeLineNV $ and $ \riverEdgeListEachLine $ \\
    $ \riverEdgeListEachLine \gets $ a river edge in $ \riverEdgeList $ \\ 
    % $ \boundingBoxInt \gets $ boolean indicating $ \nodeBoundingBox $ intersects $ \riverEdgeBoundingBoxEach $ \\
    % $ \nodeEdge \gets $ the edge for $ \node $\\
    % $ \edgeBoxInt \gets $ boolean indicating $ \nodeEdge $ intersects $ \riverEdgeListEach $ \\

    \begin{algorithmic}[1]
        \Procedure{TestRiverIntersection}{$ \node, \nodeVPSC $}

        \ForEach{$ \riverEdgeListEachLine \in \riverEdgeList $}
            \State $ \nodeBoundingBox \gets $ GetBoundingBox ($ \nodeLineNV $)
            \State $ \riverEdgeBoundingBoxEach \gets $ GetBoundingBox ($ \riverEdgeListEachLine $)
            % \State $ \boundingBoxInt \gets $ TestIntersection ($ \nodeBoundingBox, \riverEdgeBoundingBoxEach $)
            % \If{$ \boundingBoxInt = True $}

            \If{$ \nodeBoundingBox ~intersect~ \riverEdgeBoundingBoxEach = True $}
                \If{$ \nodeLineNV ~intersect~ \riverEdgeListEachLine = True $}
                    \State \Return{$ True $}
                \EndIf

@@

\subsection{Detecting Stalemates}

As the VPSC always tries to produce an optimal node layout where node distribution and translation are minimized, a node might be repeatedly translated between two positions due to river intersection and congestion, creating a stalemate situation, as shown in \figref{fig:stalemate}. In order to create space for VPSC to generate a new layout, we introduce a parameter, $ \stalemateMax $, to test for stalemate. If a node is translated between two positions for more than $ \stalemateMax $ iterations, it is considered to be in a stalemate. A stalemate triggers a corridor derivation, {a corridor is a rectangle with a width of $ \CorridorWidth $ and a length of $ \CorridorLength $, in the direction of the node $ \node $'s previous and current position, denoted $ \nodeVectorPC $ }, all enclosed nodes are then translated in the same direction and by the distance of that vector to alleviate the congestion.

@@

{
\begin{figure}[tb!]
    \centering
    \includegraphics[width=\columnwidth]{figure/corridor.png}
    \caption{When a stalemate occurs, a corridor (black rectangular) is derived based on the current (A) and previous (B) positions of the node that crossed a river, as described in \algoref{alg:derive corridor}. All nodes within the corridor are moved by $ d $ units in the direction of $ \Vector{B}{A} $, e.g, A is moved to C. }
    \label{fig:corridor}
\end{figure}
}

@@

    \textbf{Local variables:} \\
    $ \nodePrevious \gets $ the previous position of $ \node $ \\
    $ \PointP \gets $ the point extending $ \nodeVectorPC $ such that the length of $ \nodeLineNpP $ is $ \CorridorLength $\\
    $ \EdgeParallelA, \EdgeParallelB \gets $ the edges parallel to $ \nodeLineNpP $ \\
    $ corridor \gets $ a rectangle formed by $ \EdgeParallelA $ and $ \EdgeParallelB $ \\

    \begin{algorithmic}[1]
        \Procedure{ProcessStalemate}{$ \node $}
            \State TranslateNode ($ \node $)

            \State $ \PointP \gets $ \Call{DerivePoint}{$ \nodeVectorPC $, $ \CorridorLength $}

            \State $ \EdgeParallelA \gets $ \Call{DeriveParallelEdge}{$ \nodeVectorNpP $, $ \frac{\CorridorWidth}{2} $}

            \State $ \EdgeParallelB \gets $ \Call{DeriveParallelEdge}{$ \nodeVectorNpP $, $ -\frac{\CorridorWidth}{2} $}

@@

                % TODO: calculate the vector once and use it for all nodes
            \ForEach{$ \nodeInCorridorEach $ inside $ corridor $}
                \State $ \nodeInCorridorEachDx \gets \nodeInCorridorEach.x - \nodeInCorridorEach.previous.x $

@@

    \begin{algorithm}[tb!]
        \caption{Procedure to derive a point based an edge and a distance.}\label{alg:derive corridor point}
        \textbf{Input:} \\
        $ \Edge \gets $ the edge used to derive the new point \\
        $ \Distance \gets $ the distance between $ \PointP $ and $ \EdgeStart $ \\

        \textbf{Output:} \\
        A point, $ \PointP $, that is distance $ \Distance $ away from $ \EdgeStart $. \\

        \textbf{Local variables:} \\
        $ \dx, \dy \gets $ the differences in $ x, y $ for $ \EdgeStart $ and $ \EdgeEnd $ \\

        \begin{algorithmic}[1]
            \Procedure{DerivePoint}{$ \Edge $, $ \Distance $}
                \State $ \dx \gets \EdgeStart.x - \EdgeEnd.x $

                \State $ \dy \gets \EdgeStart.y - \EdgeEnd.y $

@@

    \textbf{Local variables:} \\
    $ \dx, \dy \gets $ the differences in $ x, y $ for $ \EdgeStart $ and $ \EdgeEnd $ \\
    $ \Scale \gets $ the scale of $ \frac{\Distance}{\sqrt{\dx^2 + \dy^2}} $ \\

    \begin{algorithmic}[1]
        \Procedure{DeriveParallelEdge}{$ \Edge $, $ \Distance $}
            \State $ \dx \gets \EdgeStart.x - \EdgeEnd.x $

            \State $ \dy \gets \EdgeStart.y - \EdgeEnd.y $

@@

            \State $ \EdgeParallel.start.x \gets \Scale \cdot -\dy + \EdgeStart.x $

            \State $ \EdgeParallel.start.y \gets \Scale \cdot \dx + \EdgeStart.y $

            \State $ \EdgeParallel.end.x \gets \Scale \cdot -\dy + \EdgeEnd.x $

            \State $ \EdgeParallel.end.y \gets \Scale \cdot \dx + \EdgeEnd.y $

### 2022-03-04 paper/section/related.tex
draft: update and review

\new{While we focus on rectangular and Dorling cartograms in this paper, we provide brief definitions of contiguous and non-contiguous cartograms for interested readers: Contiguous cartograms are topology-preserving that maintain connectivity with their adjacent areas but have distortion in shape. Non-contiguous cartograms sacrifice the topological connectivity to enable growing or shrinking in size to maintain their original shape \cite{dent2009Cartography}.}

\citea{nusrat2016State} defines summarize three major accuracy dimensions for cartograms: statistical, geographical, and topological. Each cartogram design may make accuracy trade-offs between dimensions, we provide an overview of these cartograms in Table \ref{table:accuracy}.

@@

\thumbnail{dorling}{
	Dorling cartograms are non-contiguous and do not preserve geography and topology. A Dorling cartogram is statistically accurate, regions are represented by circles and the data dimensions of interest are represented by the circle size \cite{dorling2011Area}. In a Demers cartogram, a variant of Dorling, squares are used instead to capture a certain level of topology \cite{cano2015Mosaic}. Dorling cartograms are unable to maintain topological accuracy as circles are often repositioned to remove overlap. \new{Here we focus on Demers cartograms as we use squares to depict regions. This style of cartogram offers the advantages that the comparisons between regions are more intuitive and the space utilization is more effcient. Demers cartograms also have the advantage of preserving a higher level of topology. }
}

\thumbnail{rectangular}{
	Rectangular cartograms are contiguous and do not preserve geographical accuracy \cite{raisz1934Rectangular, vankreveld2004Rectangular}. Depending on the variant, a rectangular cartogram may trade off between statistical and topological accuracy. \new{Statistical comparisons between regions are difficult due to the differing shapes.}
}

@@

Cartograms are a popular choice of representation in covering various topics by the media, with the goal of presenting an unbiased and complete picture of the news. The Washington Post uses cartograms to visualize the US overseas economic assistance, arm sales (Mosaic) \cite{bearak2016Everything}, the 2016 US Election (contiguous) \cite{gamio2016Election}, and the Brexit Referendum (Mosaic) \cite{taylor2016What}. The National Geographic uses contiguous and Mosaic cartograms to analyze the 2016 US Election results \cite{miller2016Election}, the same topic is also covered by the Financial Times with a Dorling cartogram \cite{stabe2016Search}. \citea{cruz2017Adapted} adapts a Dorling cartogram with both contiguous and non-contiguous cartograms to represent the gender pay gap in Portugal. \citea{sandberg2018Cartogram} reports the 2018 US midterm Election with a Mosaic cartogram, the same approach is used to cover the 2020 US Election by the New York Times \cite{thelearningnetwork2020What} and Bloomberg \cite{mccartney20202020}.

@@

Compared to previous work on Demers cartograms, we introduce dynamic topological feature, a river, aiming to improve the interpretation, readability, and topological accuracy.

>>>>

While we focus on rectangular and Dorling cartograms in this paper, we provide brief definitions of contiguous and non-contiguous cartograms for interested readers: Contiguous cartograms are topology-preserving that maintain connectivity with their adjacent neighbors but have distortion in shape. Non-contiguous cartograms sacrifice the topological connectivity with neighbors to enable growing or shrinking in size while maintaining their original shape \cite{dent2009Cartography}.

\citea{nusrat2016State} define and summarize three major accuracy dimensions for cartograms: statistical, geographical, and topological. Each cartogram design may make accuracy trade-offs between dimensions. We provide an overview of these cartograms in Table \ref{table:accuracy}.

@@

\thumbnail{dorling}{
	Dorling cartograms are non-contiguous and do not preserve geography and topology. A Dorling cartogram is statistically accurate, regions are represented by circles and the data dimensions of interest are represented by the circle size \cite{dorling2011Area}. In a Demers cartogram, a variant of Dorling, squares are used instead to capture a certain level of topology \cite{cano2015Mosaic}. Dorling cartograms are unable to maintain topological accuracy as circles are often repositioned to remove overlap. Here we focus on Demers cartograms as we use squares to depict regions. This style of cartogram offers the advantages that the comparisons between regions are intuitive and screen space utilization is more efficient. Demers cartograms also have the advantage of preserving a higher level of topology. 
}

\thumbnail{rectangular}{
	Rectangular cartograms are contiguous and do not preserve geographical accuracy \cite{raisz1934Rectangular, vankreveld2004Rectangular}. Depending on the variant, a rectangular cartogram may trade off between statistical and topological accuracy.
}

@@

Cartograms are a popular choice of representation in covering various topics by the media, with the goal of presenting an unbiased and complete picture of the news. The Washington Post uses cartograms to visualize the US overseas economic assistance, arm sales (Mosaic) \cite{bearak2016Everything}, the 2016 US Election (contiguous) \cite{gamio2016Election}, and the Brexit Referendum (Mosaic) \cite{taylor2016What}. National Geographic uses contiguous and Mosaic cartograms to analyze the 2016 US Election results \cite{miller2016Election}, the same topic is also covered by the Financial Times with a Dorling cartogram \cite{stabe2016Search}. \citea{cruz2017Adapted} adapts a Dorling cartogram with both contiguous and non-contiguous cartograms to represent the gender pay gap in Portugal. \citea{sandberg2018Cartogram} reports the 2018 US midterm Election with a Mosaic cartogram, the same approach is used to cover the 2020 US Election by the New York Times \cite{thelearningnetwork2020What} and Bloomberg \cite{mccartney20202020}.

@@

Our work extends Demers cartograms. We introduce dynamic topological feature, a river, aiming to improve the interpretation, readability, and geographical accuracy.
